Modelling low Mach number stellar hydrodynamics with MAESTROeX

### 2019-10-02 astronum_2019_MAESTROeX/astronum-maestroex.tex
updated abstract; some other minor edits

\begin{abstract}
Convective flows in the interiors of stars are challenging to model using conventional compressible hydrodynamics codes due to their low Mach number, which severely limits the timestep that can be taken in simulations. \maestroex\ is an open source low Mach number stellar hydrodynamics code that allows much larger timesteps to be taken, therefore enabling systems to be modelled for much longer periods of time. This is particularly important for the problem of convection in the cores of rotating massive stars prior to core collapse. To fully capture the dynamics, it is necessary to model these systems in 3d at high resolution over many rotational periods, but this has been too computationally expensive for existing, conventional schemes.

We shall present an overview of \maestroex's current capabilities, describe ongoing work to incorporate the effects of rotation and discuss how we are optimizing the code to run on GPUs. 
\end{abstract}

@@

The massive stars that we are interested in modelling are believed to be relatively slow rotators\MarginPar{insert some citation/quantify this}, so it is likely that one of the simpler approaches outlined above should be sufficient. However, for modelling e.g.~type I X-ray bursts, it may be that a more sophisticated approach is needed. Type I X-ray bursts are produced by thermonuclear deflagrations in the liquid surface layers of fast rotating neutron stars. These stars typically rotate at frequencies of $\sim$300\textendash600 Hz, a significant fraction of the stars' breakup velocity. This produces non-negligible deformation of the star about the equator, and may cause surface flows with a significant dependence on the meriodional angle. To properly capture the effects of rotation in such a system, it may therefore be necessary to use a 2d base state, with the base state pressure a function of both the radius and the meriodional angle $p_0(r)\rightarrow p_0(r, \theta)$.

% \begin{itemize}
% \item Ignore centrifugal force \textemdash~for slow rotation can neglect centrifugal, only include Coriolis force
% \item Introduce new pressure to balance centrifugal potential $$\rho \nabla \Phi_{\text{eff}} = \nabla p_0 + \nabla p_1 = \rho \nabla \phi + \frac{1}{2}\rho \nabla\Omega^2 r^2$$
% \item Rewrite base state in terms of equipotentials $$p_0(r) \rightarrow p_0(\Phi_{\text{eff}}) $$
% \item 2d base state $$p_0(r)  \rightarrow p_0(r, \theta)$$
% \end{itemize}

@@

\MarginPar{I feel like someone who knows what they're talking about should write/seriously edit this section}
The latest generation of supercomputers to come online (e.g.~Summit, soon Perlmutter and Frontier) rely heavily on GPUs in order to achieve high performance while keeping energy consumption to a minimum (GPUs consume considerably less power per flop than CPUs). This trend is set to continue: supercomputers are moving to new architectures in order to increase their performance, rather than simply adding more and more CPUs. In order for HPC codes to be able to exploit these latest machines, it is therefore becoming increasingly necessary to port codes to run on GPUs.

We have begun porting \maestroex~to GPUs, leveraging the GPU capabilities built into \amrex~and following the lessons learned porting \castro~to GPUs. In the \amrex~framework, routines can be ported to GPUs by simply inserting a few macros into the code. The boilerplate code to required to compile this for GPU is then generated by a custom preprocessor prior to compilation. CUDA managed memory is used to take care of data transfers between the CPU and GPU, so no explicit copies of data to/from the CPU and GPU are required, significantly reducing the extra code required to port the code. One of the key emphases of the \amrex~approach to GPUs is to maintain performance portability: porting code to GPUs should not come at the expense of the code's performance on CPU-only machines. Because GPU-specific code is created by the preprocessor prior to compilation, it is not necessary to maintain separate versions of the code for different architectures. By changing the compiler flags, the same bit of code can be compiled for serial, MPI-only, OpenMP-only, MPI+OpenMP, GPU-only or MPI+GPU.   
In order to achieve the best performance on the GPU, some additional optimization is required. However, it has been found that these optimizations often improve the performance when the code is run on the CPU only as well. 
More details to appear in a forthcoming \amrexastro~GPU paper.

So far we have ported most of the source terms and hydrodynamics in \maestroex~to run on GPUs. We have found that one of the challenges moving forwards is dealing with the 1d base state. Currently, all AMR subgrids have a copy of the entire 1d base state. This is not a problem for small problems where the 1d base state does not contain many cells compared to the number of cells in a single subgrid, however as the problem size and/or resolution gets larger, the size of the 1d base state grows to the point that it is now significantly larger than the size of a single subgrid. Compared to CPUs, GPUs typically have less memory. For larger problems, it is therefore not possible to store the entire base state on the GPU. This significantly reduces the performance of the GPU parts of the code, as it will increase the amount of data will be copied back and forth between the CPU and GPU (a slow process). To alleviate this problem, we are looking at ways to reduce the amount of 1d base state data that needs to be stored on the GPU at any one time. For example, in some routines we map the 1d base state to the 3d grid \textendash~for these, we can move this operation to the CPU so that the GPU does not require the 1d base state. It may be necessary for us to develop a way to split up the 1d base state, so that subgrids only receive the sections of the base state necessary for their calculations (rather than the base state for the entire problem domain).

@@

    \caption{Comparison of the performance of functions on the CPU vs the GPU. The plot shows the average execution time for several functions, recorded for a run of the 3d \texttt{reacting\_bubble} problem on Summit. Both tests were performed using a single node, with the CPU test using 42 MPI processes, each with 4 OpenMP threads, and the GPU test using 6 MPI processes, each process associated with a single GPU. Note that the $x$-axis scale is logarithmic.}

@@

We're currently working on implementing rotation in \maestroex. For modelling the interiors of massive stars (which rotate relatively slowly), we may be able to get away with neglecting the centrifugal force and simply add the Coriolis force as a source term. However, this will not be sufficient for faster rotating systems (e.g.~for modelling Type I X-ray bursts on the surfaces of millisecond pulsars), so we are also exploring other methods including modelling the centrifugal force as an effective pressure term, rewriting the base state as a function of the effective gravitational potential and using a 2d base state which is also a function of the meriodional angle.

>>>>

\begin{abstract}
Modelling long-time convective flows in the interiors of stars is extremeley challenging using conventional compressible hydrodynamics codes due to the acoustic timestep limitation.
Many of these flows are in the low Mach number regime, which allows us to exploit the relationship between acoustic and advective time scales to develop a more computationally efficient approach.
\maestroex\ is an open source low Mach number stellar hydrodynamics code that allows much larger timesteps to be taken, therefore enabling systems to be modelled for much longer periods of time. This is particularly important for the problem of convection in the cores of rotating massive stars prior to core collapse. To fully capture the dynamics, it is necessary to model these systems in three dimensions at high resolution over many rotational periods.  We present an overview of \maestroex's current capabilities, describe ongoing work to incorporate the effects of rotation and discuss how we are optimizing the code to run on GPUs. 
\end{abstract}

@@

The massive stars that we are interested in modelling are believed to be relatively slow rotators\MarginPar{insert some citation/quantify this}, so it is likely that one of the simpler approaches outlined above should be sufficient. However, for modelling e.g.~type I X-ray bursts, it may be that a more sophisticated approach is needed. Type I X-ray bursts are produced by thermonuclear deflagrations in the liquid surface layers of fast rotating neutron stars. These stars typically rotate at frequencies of $\sim$300\textendash600 Hz, a significant fraction of the stars' breakup velocity. This produces non-negligible deformation of the star about the equator, and may cause surface flows with a significant dependence on the meriodional angle. To properly capture the effects of rotation in such a system, it may therefore be necessary to use a two-dimensional base state, with the base state pressure a function of both the radius and the meriodional angle $p_0(r)\rightarrow p_0(r, \theta)$.

% \begin{itemize}
% \item Ignore centrifugal force \textemdash~for slow rotation can neglect centrifugal, only include Coriolis force
% \item Introduce new pressure to balance centrifugal potential $$\rho \nabla \Phi_{\text{eff}} = \nabla p_0 + \nabla p_1 = \rho \nabla \phi + \frac{1}{2}\rho \nabla\Omega^2 r^2$$
% \item Rewrite base state in terms of equipotentials $$p_0(r) \rightarrow p_0(\Phi_{\text{eff}}) $$
% \item two-dimensional base state $$p_0(r)  \rightarrow p_0(r, \theta)$$
% \end{itemize}

@@

\MarginPar{I feel like someone who knows what they're talking about should write/seriously edit this section}
The latest generation of supercomputers to come online (e.g.~Summit, soon Perlmutter and Frontier) rely heavily on GPUs in order to achieve high performance while keeping energy consumption to a minimum (GPUs consume considerably less power per flop than CPUs). This trend is set to continue: supercomputers are moving to new architectures in order to increase performance rather than simply adding more and more CPUs. In order for HPC codes to be able to exploit these latest machines, it is therefore becoming increasingly necessary to port codes to run on GPUs.

We have begun porting \maestroex~to GPUs, leveraging the GPU capabilities built into \amrex~and following the lessons learned porting \castro~to GPUs. In the \amrex~framework, routines can be ported to GPUs by simply inserting a few macros into the code. The boilerplate code to required to compile this for GPU is then generated by a custom preprocessor prior to compilation. CUDA managed memory is used to take care of data transfers between the CPU and GPU, so no explicit copies of data to/from the CPU and GPU are required, significantly reducing the extra code required to port the code. One of the key emphases of the \amrex~approach to GPUs is to maintain performance portability: porting code to GPUs should not come at the expense of the code's performance on CPU-only machines. Because GPU-specific code is created by the preprocessor prior to compilation, it is not necessary to maintain separate versions of the code for different architectures. By changing the compiler flags, the same bit of code can be compiled for serial, MPI-only, OpenMP-only, MPI+OpenMP, GPU-only or MPI+GPU.   
In order to achieve the best performance on the GPU, we require some additional optimization of the computational kernals. However, we find that these optimizations often improve the performance when the code is run on the CPU only as well. 
More details to appear in a forthcoming \amrexastro~GPU paper.

So far we have ported most of the source terms and hydrodynamics in \maestroex~to run on GPUs. GPU support for the linear solvers is becoming increasingly available from the \amrex~team.  We have found that one of the challenges moving forwards is dealing with the one-dimensional base state. Currently, each MPI process holds a copy of the entire base state. This is not a problem for small problems where the base state does not contain many cells compared to the number of cells in a single subgrid, however as the problem size and/or resolution gets larger, the size of the base state grows to the point that it is now significantly larger than the size of a single subgrid. Compared to CPUs, GPUs typically have less memory. For larger problems, it is therefore not possible to store the entire base state on the GPU. This significantly reduces the performance of the GPU parts of the code, as it will increase the amount of data will be copied back and forth between the CPU and GPU (a slow process). To alleviate this problem, we are looking at ways to reduce the amount of base state data that needs to be stored on the GPU at any one time. For example, in some routines we map the base state to the three-dimensional grid \textendash~for these, we can move this operation to the CPU so that the GPU does not require the base state. It may be necessary for us to develop a way to split up the base state, so that subgrids only receive the sections of the base state necessary for their calculations (rather than the base state for the entire problem domain).

@@

    \caption{Comparison of the performance of functions on the CPU vs the GPU. The plot shows the average execution time for several functions, recorded for a run of the three-dimensional \texttt{reacting\_bubble} problem on Summit. Both tests were performed using a single node, with the CPU test using 42 MPI processes, each with 4 OpenMP threads, and the GPU test using 6 MPI processes, each process associated with a single GPU. Note that the $x$-axis scale is logarithmic.}

@@

We're currently working on implementing rotation in \maestroex. For modelling the interiors of massive stars (which rotate relatively slowly), we may be able to get away with neglecting the centrifugal force and simply add the Coriolis force as a source term. However, this will not be sufficient for faster rotating systems (e.g.~for modelling Type I X-ray bursts on the surfaces of millisecond pulsars), so we are also exploring other methods including modelling the centrifugal force as an effective pressure term, rewriting the base state as a function of the effective gravitational potential and using a two-dimensional base state which is also a function of the meriodional angle.
